Open Data Quality Is Poor but Slowly Improving

### 2021-01-30 outputs/paper/paper.Rmd
final edits

More than a year since the Data Quality Score was first used, we can begin to see patterns: scoring is irregular and quality is poor overall, but there is some improvement over time. This report will first consider the scoring data relative to time to consider how different scoring dimensions perform relative to one another, finding that Metadata and Freshness scores are poor. Second, it will demonstrate that scoring has taken place over uneven intervals of time. Third, it will hypothesize that increases in average Quality scores may be associated with increased in number of packages. Fourth, it will compare the grade levels given in January 2021 to those given six months prior and find that the number of higher-grade packages is increasing. Finally, it will discuss the limits of the Data Quality Score, in terms of bias and in terms of measuring the societal value of the portal.

@@

When scoring is done, all possible packages on the portal are scored at the same time. In theory, this should reduce bias, as the sample is the entire population. However, many of the packages on the portal are not eligible for scoring. Data files, such as , are not scored. Currently, only data that is in the CKAN Datastore API is scored, due to ease of access for scoring [@HernandezMedium2]. This introduces an obvious source of bias: files that are not available through the API are often large (zip files) or non-ideal formats (Excel or PDF), meaning that the lower-quality packages may have been disproportionately excluded from the scoring. For this reason, the average scores can only be considered to reflect a specific subcategory of the whole data catalogue. Another issues is that Read Me files are scored and weighted exactly the same as data resources, despite having very different values and qualities [@metaopendata].

@@

```{r facetedscorestime, fig.cap = "Scores over time", fig.width = 8, fig.height = 8, echo = FALSE, message = FALSE, warning = FALSE, out.width = '75%', fig.align = "center"}
cleaned_data %>%
  # Create a pivot table so the different types of scores can easily be faceted:
  tidyr::pivot_longer(
    cols = c("quality", "completeness", "freshness", "metadata", "usability", "accessibility"),
    names_to = "metric",
    values_to = "scored"
    ) %>%

@@

  # Make a point plot to compare scores over time:
  ggplot(aes(x = date_scored, y = scored)) +
  geom_point(alpha = 0.05, color = "darkturquoise", size = .5) +
  geom_smooth(colour = "magenta3", size = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Date Scored",
       y = "Grade",
       title = "Scores Are Relatively Stable Over Time") +

  # Correctly format the dates for easy readability:
  scale_x_datetime(breaks = c(as.POSIXct("2020-01-01 00:01 EST"), 
                              as.POSIXct("2020-03-01 01:01 EST"),
                              as.POSIXct("2020-05-01 01:01 EST"),
                              as.POSIXct("2020-07-01 01:01 EST"),
                              as.POSIXct("2020-09-01 01:01 EST"),
                              as.POSIXct("2020-11-01 01:01 EST"),
                              as.POSIXct("2021-01-01 01:01 EST")),
                    minor_breaks = NULL,
                    labels = date_format("%b\n%Y", tz="EST"),  # Format the x-axis scale to show month names
                   # set x-axis limits 
                   limits = as.POSIXct(strptime(c("2019-12-16 00:00", "2021-01-25 00:00"), 
                                                 format = "%Y-%m-%d %H:%M"))
                   ) +

@@

```{r monthbreakdown, echo = FALSE, message = FALSE, warning = FALSE}
cleaned_data %>%

@@

  # Summarise counts and means for each month:
  summarise(scored = n_distinct(package),
            scores = n_distinct(date_scored),
            mean_accessibility = mean(accessibility),
            mean_completeness = mean(completeness),
            mean_freshness = mean(freshness),
            mean_metadata = mean(metadata),
            mean_usability = mean(usability),
            mean_quality = mean(quality)
  ) %>%

  # Put the results in a table:
  knitr::kable(digits = 3,
               col.names = c("Year and Month", "Unique Packages Scored",
                             "Times the Portal was Scored", "Accessibility",
                             "Completeness", "Freshness", "Metadata",
                             "Usability", "Quality"),
               booktabs = TRUE,
               escape = FALSE,
               caption = "Scoring by month"
  ) %>%

  # Style the table:
  add_header_above(c("", "Counts" = 2, "Average Scores" = 6)) %>%
  column_spec(1, width = "5em") %>%
  column_spec(2, width = "5em") %>%
  column_spec(3, width = "5em") %>%
  column_spec(4, width = "5em") %>%
  column_spec(5, width = "5em") %>%
  column_spec(6, width = "5em") %>%
  column_spec(7, width = "5em") %>%
  column_spec(8, width = "5em") %>%
  column_spec(9, width = "5em") %>%
  pack_rows("Manual Scoring", 1, 3) %>%
  pack_rows("Scheduled Scoring", 4, 10) %>%
  kableExtra::kable_styling(latex_options = c("scale_down", "hold_position"))

@@

  scale_x_datetime(breaks = c(as.POSIXct("2020-01-01 00:01 EST"), # Set breaks on the first of the month
                              as.POSIXct("2020-02-01 01:01 EST"),
                              as.POSIXct("2020-03-01 01:01 EST"),
                              as.POSIXct("2020-04-01 01:01 EST"),
                              as.POSIXct("2020-05-01 01:01 EST"),
                              as.POSIXct("2020-06-01 01:01 EST"),
                              as.POSIXct("2020-07-01 01:01 EST"),
                              as.POSIXct("2020-08-01 01:01 EST"),
                              as.POSIXct("2020-09-01 01:01 EST"),
                              as.POSIXct("2020-10-01 01:01 EST"),
                              as.POSIXct("2020-11-01 01:01 EST"),
                              as.POSIXct("2020-12-01 01:01 EST"),
                              as.POSIXct("2021-01-01 01:01 EST")),
                    minor_breaks = NULL,  # Remove minor break lines
                    labels = date_format("%b\n%Y",tz="EST"),  # Format the x-axis scale to show month names
                    limits = as.POSIXct(strptime(c("2019-12-16 00:00", "2021-01-25 00:00"),  # x-axis limits
                    format = "%Y-%m-%d %H:%M"))) +
  labs(color = "Average \nQuality \nscore \nacross all \npackages",
       x = "Date recorded",
       y = "Number of packages scored",
       title = "Scoring Frequency Increased Over the Beta Testing Period") + 
  theme_light()
```

@@

```{r silverscore, fig.cap = "The Catalogue Quality Scores package receives a silver grade", echo = FALSE, message = FALSE, warning = FALSE, out.width = '100%'}

>>>>

More than a year since the Data Quality Score was first used, we can begin to see patterns: scoring is irregular and quality is poor overall, but there is some improvement over time. This report will first consider the scoring data relative to time to consider how different scoring dimensions perform relative to one another, finding that Metadata and Freshness scores are poor. Second, it will demonstrate that scoring has taken place over uneven intervals of time. Third, it will hypothesize that increases in average Quality scores may be associated with increases in number of packages. Fourth, it will compare the grade levels given in January 2021 to those given six months prior and find that the number of higher-grade packages is increasing. Finally, it will discuss the limits of the Data Quality Score, in terms of bias and in terms of measuring the societal value of the portal.

@@

When scoring is done, all possible packages on the portal are scored at the same time. In theory, this should reduce bias, as the sample is the entire population. However, many of the packages on the portal are not eligible for scoring. Currently, only data that is in the CKAN Datastore API is scored, due to ease of access for scoring [@HernandezMedium2]. This introduces an obvious source of bias: files that are not available through the API are often large (zip files) or non-ideal formats (Excel or PDF), meaning that the lower-quality packages may have been disproportionately excluded from the scoring. For this reason, the average scores can only be considered to reflect a specific subcategory of the whole data catalogue. Another issue is that Read Me files are scored and weighted exactly the same as data resources, despite having very different values and qualities [@metaopendata].

@@

```{r facetedscorestime, fig.cap = "Scores over time", fig.width = 8, fig.height = 8, echo = FALSE, message = FALSE, warning = FALSE, out.width = '80%', fig.align = "center"}
cleaned_data %>%
  # Create a pivot table so the different types of scores can easily be faceted:
  tidyr::pivot_longer(
    cols = c("quality", "completeness", "freshness", "metadata", "usability", "accessibility"),
    names_to = "metric",
    values_to = "scored"
    ) %>%

@@

  # Make a point plot to compare scores over time:
  ggplot(aes(x = date_scored, y = scored)) +
  geom_point(alpha = 0.05, color = "darkturquoise", size = .5) +
  geom_smooth(colour = "magenta3", size = 1) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Date scored",
       y = "Grade",
       title = "Scores are relatively stable over time") +

  # Correctly format the dates for easy readability:
  scale_x_datetime(limits = as.POSIXct(strptime(c("2019-12-16 10:00:00", "2021-01-25 10:00:00"), 
                                                 format = "%Y-%m-%d %H:%M:%S")),
                   expand = c(0, 0),  # keep the sides of the plot snug to the limits
                   labels = date_format("%b\n%Y"),  # Format the x-axis scale to show month names
                   breaks = c(as.POSIXct("2020-01-01 01:00:00"),  # Set breaks on the first of the month
                              as.POSIXct("2020-03-01 01:00:00"),
                              as.POSIXct("2020-05-01 01:00:00"),
                              as.POSIXct("2020-07-01 01:00:00"),
                              as.POSIXct("2020-09-01 01:00:00"),
                              as.POSIXct("2020-11-01 01:00:00"),
                              as.POSIXct("2021-01-01 01:00:00")),
                    minor_breaks = NULL,  # Remove minor break lines
                   ) +

@@

```{r monthbreakdown, echo = FALSE, message = FALSE, warning = FALSE}
df <- cleaned_data %>%

@@

  # Summarise counts and means for each month:
  summarise(scored = n_distinct(package),
            scores = n_distinct(date_scored),
            mean_accessibility = mean(accessibility),
            mean_completeness = mean(completeness),
            mean_freshness = mean(freshness),
            mean_metadata = mean(metadata),
            mean_usability = mean(usability),
            mean_quality = mean(quality)
  ) %>%
  # Tidy up names so they are better for the table display:
  rename("Unique Packages Scored" = scored,
         "Times the Portal was Scored" = scores, 
         "Accessibility" = mean_accessibility,
         "Completeness" = mean_completeness, 
         "Freshness" = mean_freshness, 
         "Metadata" = mean_metadata,
         "Usability" = mean_usability, 
         "Quality" = mean_quality) %>%
  mutate_if(is.numeric, round, 3) # round score averages to 3 decimal places

# transpose the tibble: from https://stackoverflow.com/questions/42790219/how-do-i-transpose-a-tibble-in-r
as_tibble(cbind(nms = names(df), t(df)))  %>%
  slice(2:9) %>% #removing the top row, which contains the dates--these are replaced below as column names.

  # Put the results in a table:
  knitr::kable(col.names = c("",
                             "Dec 2019",
                             "Feb 2020",
                             "Mar 2020",
                             "Jul 2020",
                             "Aug 2020",
                             "Sept 2020",
                             "Oct 2020",
                             "Nov 2020",
                             "Dec 2020",
                             "Jan 2021"),
               booktabs = TRUE,
               escape = FALSE,
               caption = "Scoring by month"
  ) %>%

  # Style the table:
  add_header_above(c("", "Manual Scoring" = 3, "Scheduled Scoring" = 7), align = "l") %>%
  column_spec(1, width = "14em") %>%
  column_spec(2, width = "3em") %>%
  column_spec(3, width = "3em") %>%
  column_spec(4, width = "3em") %>%
  column_spec(5, width = "3em") %>%
  column_spec(6, width = "3em") %>%
  column_spec(7, width = "3em") %>%
  column_spec(8, width = "3em") %>%
  column_spec(9, width = "3em") %>%
  column_spec(10, width = "3em") %>%
  column_spec(11, width = "3em") %>%
  pack_rows("Counts", 1, 2) %>%
  pack_rows("Average Scores", 3, 8) %>%
  kableExtra::kable_styling(latex_options = c("scale_down", "hold_position"))
```

@@

  scale_x_datetime(limits = as.POSIXct(strptime(c("2019-12-16 10:00:00", "2021-01-25 10:00:00"), 
                                                 format = "%Y-%m-%d %H:%M:%S")),
                   expand = c(0, 0),  # keep the sides of the plot snug to the limits
                   labels = date_format("%b\n%Y"),   # Format the x-axis scale to show month names
                   breaks = c(as.POSIXct("2020-01-01 01:00:00"),  # Set breaks on the first of the month
                              as.POSIXct("2020-02-01 01:00:00"),
                              as.POSIXct("2020-03-01 01:00:00"),
                              as.POSIXct("2020-04-01 01:00:00"),
                              as.POSIXct("2020-05-01 01:00:00"),
                              as.POSIXct("2020-06-01 01:00:00"),
                              as.POSIXct("2020-07-01 01:00:00"),
                              as.POSIXct("2020-08-01 01:00:00"),
                              as.POSIXct("2020-09-01 01:00:00"),
                              as.POSIXct("2020-10-01 01:00:00"),
                              as.POSIXct("2020-11-01 01:00:00"),
                              as.POSIXct("2020-12-01 01:00:00"),
                              as.POSIXct("2021-01-01 01:00:00")),
                    minor_breaks = NULL,  # Remove minor break lines
                   ) +
  labs(color = "Average \nQuality \nscore \nacross all \npackages",
       x = "Date recorded",
       y = "Number of packages scored",
       title = "Scoring frequency increased over the beta testing period") + 
  theme_light()
```

@@

```{r silverscore, fig.cap = "The 'Catalogue quality scores' package receives a silver grade", echo = FALSE, message = FALSE, warning = FALSE, out.width = '100%'}
